---
title: "PA1_Template"
author: "Srikar Murali"
date: "October 14, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(stringr)
library(xts)

```

## Load Data


```{r load}
data <- read.csv('C:/Users/srika/Documents/R/John Hopkins/Reproducible Research/activity.csv', header = TRUE)
data2 <- na.omit(data)

data2$date <- as.Date(data2$date)



```

## Mean total number of steps

You can also embed plots, for example:

```{r mean1}
dailySteps <- data2 %>%
  group_by(date) %>%
  summarise(TotalSteps = sum(steps))
dailySteps
stepHist <- ggplot(dailySteps, aes(TotalSteps)) +
  geom_histogram(fill = 'blue') +
  xlab('Total Number of steps per day') +
  ylab('Count') +
  ggtitle('Histogram of steps by day')
stepHist

averageSteps <- mean(dailySteps$TotalSteps)
#Average number of steps
averageSteps

medianSteps <- median(dailySteps$TotalSteps)
#Median number of steps
medianSteps

stepsTSeries <- xts(dailySteps$TotalSteps, order.by = dailySteps$date)

```

The average number of steps is 10766.19. The median number of steps is 10765.

```{r average daily pattern}
intervalTSeries <- data2 %>%
  group_by(interval) %>%
  summarise(AvgSteps = mean(steps))
intervalTSeries

plot(intervalTSeries$interval, intervalTSeries$AvgSteps,
     type = 'l',
     xlab = 'Interval',
     ylab = 'Average Steps',
     main = 'Average Steps by Interval')
maxSteps <- intervalTSeries$interval[which.max(intervalTSeries$AvgSteps)]
maxSteps
```

The 835th interval had the max steps.

## Remove Na's
```{r NA}

nas <- sum(is.na(data$steps))
nas

filledData <- data

meansteps <- mean(data2$steps)
meansteps
filledData$steps[is.na(filledData$steps)] <- meansteps

filledDataNas <- sum(is.na(filledData$steps))
filledDataNas

dailySteps2 <- filledData %>%
  group_by(date) %>%
  summarise(TotalSteps = sum(steps))
dailySteps2

stepHist2 <- ggplot(dailySteps2, aes(TotalSteps)) +
  geom_histogram(fill = 'blue') +
  xlab('Total Number of steps per day') +
  ylab('Count') +
  ggtitle('Histogram of steps by day')
stepHist2

medianSteps2 <- median(dailySteps2$TotalSteps)
medianSteps2

meanSteps2 <- mean(dailySteps2$TotalSteps)
meanSteps2

```

There are 2304 NA's. The NA's were replaced by the mean. The mean stayed the same while the median became the mean, since all of the NA's were replaced by the mean. The histogram became slightly different, with it being more bunched in the middle since all of the NA's are now the mean.


## Weekends and Weekdays
```{r weekdays}
filledData$date <- as.Date(filledData$date)
filledData$Day <- weekdays(filledData$date)


filledData$dtype <- 'weekday'
filledData$dtype[filledData$Day %in% c('Saturday', 'Sunday')] <- 'weekend'

avgDay <- filledData %>%
  group_by(dtype, interval) %>%
  summarise(AvgSteps = mean(steps))
avgDay

t <- ggplot(avgDay, aes(interval, AvgSteps)) +
  geom_line() + 
  xlab('Interval') +
  ylab('Average Step Count') +
  ggtitle('Average steps by Weekend vs Weekday') +
  facet_wrap(~dtype)
t

```

Activity is earlier on the weekdays as people have to get up early to go to work or school or some other place.


