---
title: "Impact of severe weather events on public health and the economy in the United States"
author: "Srikar Murali"
date: "October 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```


```{r libraries, echo=FALSE,message=FALSE,warning=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)


```

## Load Data

```{r start}


data <- read.csv('C:/Users/srika/Documents/R/John Hopkins/Reproducible Research/repdata%2Fdata%2FStormData.csv')
str(data)
```


## Process Data

There are 7 variables we are interested regarding the two questions. They are:-

EVTYPE as a measure of event type (e.g. tornado, flood, etc.)
FATALITIES as a measure of harm to human health
INJURIES as a measure of harm to human health
PROPDMG as a measure of property damage and hence economic damage in USD
PROPDMGEXP as a measure of magnitude of property damage (e.g. thousands, millions USD, etc.)
CROPDMG as a measure of crop damage and hence economic damage in USD
CROPDMGEXP as a measure of magnitude of crop damage (e.g. thousands, millions USD, etc.)

```{r Process}
requiredData <- select(data, BGN_DATE,EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
requiredData$byYear <- as.numeric(format(as.Date(requiredData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))

```


## Most harm to humans and economy

```{r Human Damage}
stormHist <- ggplot(requiredData, aes(byYear)) +
  geom_histogram(bins = 25, fill = 'blue') +
  ggtitle('Weather Fatilities by Year')
stormHist

fatalities <- requiredData %>%
  group_by(EVTYPE) %>%
  summarise(TotalFatilities = sum(FATALITIES)) %>%
  arrange(desc(TotalFatilities)) %>%
  top_n(10)
eventFatalities <- ggplot(fatalities, aes(x = EVTYPE, y = TotalFatilities)) +
  geom_bar(stat = 'identity',fill = 'purple') +
  xlab('Event Type') +
  ggtitle('Fatilities by Event')
eventFatalities

injuries <- requiredData %>%
  group_by(EVTYPE) %>%
  summarise(TotalInjured = sum(INJURIES)) %>%
  arrange(desc(TotalInjured)) %>%
  top_n(10)
eventInjured <- ggplot(injuries, aes(x = EVTYPE, y = TotalInjured)) +
  geom_bar(stat = 'identity', fill = 'green') +
  xlab('Event Type') +
  ggtitle('People Injured by Event')
eventInjured



```

As you can see from the analysis, TORNADO caused the most fatalities and most injuries. FLOOD caused the most property damage. DROUGHT caused the most crop damange, while FLOOD caused the most overall economic damage.


## Most Harm to Crops and Property

```{r Crop/Prop Damage}
unique(requiredData$PROPDMGEXP)
unique(requiredData$CROPDMGEXP)


requiredData$PROPDMGEXP <- as.character(requiredData$PROPDMGEXP)
requiredData$PROPDMGEXP <- gsub("\\-|\\+|\\?","0",requiredData$PROPDMGEXP)
requiredData$PROPDMGEXP <- gsub('B|b', '9', requiredData$PROPDMGEXP)
requiredData$PROPDMGEXP <- gsub('M|m', '6', requiredData$PROPDMGEXP)
requiredData$PROPDMGEXP <- gsub('K|k', '3', requiredData$PROPDMGEXP)
requiredData$PROPDMGEXP <- gsub('H|h', '2', requiredData$PROPDMGEXP)
requiredData$PROPDMGEXP <- as.numeric(requiredData$PROPDMGEXP)
requiredData$PROPDMGEXP[is.na(requiredData$PROPDMGEXP)] <- 0
requiredData$TrueDamage <- requiredData$PROPDMG * 10^requiredData$PROPDMGEXP

propertyDamage <- aggregate(TrueDamage ~ EVTYPE, data = requiredData, sum)
propertyDamage <- propertyDamage[order(-propertyDamage$TrueDamage),]
propertyDamage <- propertyDamage[1:10,]


requiredData$CROPDMGEXP <- as.character(requiredData$CROPDMGEXP)
requiredData$CROPDMGEXP <- gsub("\\-|\\+|\\?","0",requiredData$CROPDMGEXP)
requiredData$CROPDMGEXP <- gsub('B|b', '9', requiredData$CROPDMGEXP)
requiredData$CROPDMGEXP <- gsub('M|m', '6', requiredData$CROPDMGEXP)
requiredData$CROPDMGEXP <- gsub('K|k', '3', requiredData$CROPDMGEXP)
requiredData$CROPDMGEXP <- gsub('H|h', '2', requiredData$CROPDMGEXP)
requiredData$CROPDMGEXP <- as.numeric(requiredData$CROPDMGEXP)
requiredData$CROPDMGEXP[is.na(requiredData$CROPDMGEXP)] <- 0
requiredData$TrueCropDamage <- requiredData$CROPDMG * 10^requiredData$CROPDMGEXP

cropDamage <- aggregate(TrueCropDamage ~ EVTYPE, data = requiredData, sum)
cropDamage <- cropDamage[order(-cropDamage$TrueCropDamage),]
cropDamage <- cropDamage[1:10,]

damageCombined <- aggregate(TrueDamage + TrueCropDamage ~ EVTYPE, data = requiredData, sum)
names(damageCombined)[2] <- 'Total_Damage'
damageCombined <- damageCombined %>%
  arrange(desc(Total_Damage)) %>%
  top_n(10)
propertyDamage
cropDamage
damageCombined  


par(mfrow = c(1,3))
DamagetoCrops <- ggplot(cropDamage, aes(x = EVTYPE, y = TrueCropDamage)) +
  geom_bar(stat = 'identity', fill = 'red') +
  xlab('Event Type') +
  ggtitle('Crop Damage by Event Type')
damagetoProperty <- ggplot(propertyDamage, aes(x = EVTYPE, y = TrueDamage)) +
  geom_bar(stat = 'identity', fill = 'red') +
  xlab('Event Type') +
  ggtitle('Property Damage by Event Type')

fullDamage <- ggplot(damageCombined, aes(x = EVTYPE, y = Total_Damage)) +
  geom_bar(stat = 'identity', fill = 'red') +
  xlab('Event Type') +
  ggtitle('Total Damage by Event Type')

DamagetoCrops
damagetoProperty
fullDamage


```

As you can see from the analysis, TORNADO caused the most fatalities and most injuries. FLOOD caused the most property damage. DROUGHT caused the most crop damange, while FLOOD caused the most overall economic damage.